Joint estimation of phylogenies and ancestral states is not consistent

### 2018-01-02 appendix.tex
consistent Pr() notation and update table

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
$$
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge \emph{fidelity} for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
%EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
%das: Does L540 not help with this? Are you thinking we should add another column with "Notation" or something?
\caption{Site pattern probabilities $\siteSplit_j$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}(a)---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
As described above we cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
$$
P(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
$$
P(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

@@

To compute the likelihood of observing a set of data, we need $P(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.

>>>>

For the $i$th factor of \eqref{eq:full_likelihood},
$$
\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot \Pr(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$ and
\begin{align*}
    \Pr(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral splits for that particular site split, topology and set of branch lengths, and $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective, and from this we have
\begin{align*}
\max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= \Pr(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
$$
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
&\qquad = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{split}
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)
$$
and the partial log-likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\end{equation}

@@

We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge \emph{fidelity} for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\emptyset} := \Pr(\siteSplitRV=\emptyset \mid \tau_1,t),
$$
with similar definitions for the remaining generating probabilities.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
%EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\subsection*{Example}
We follow with an expository example computing these probabilities and likelihoods.
Consider the fixed, binary four-tip tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}(a)---this is commonly known as the ``Farris zone'' topology.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{3\}$ and internal node $\{2\}$ connected to leaves $\{2\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
As described above we cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum occurs at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$, defined via $\ancestralSplitPartition_j(\tau, t)$ for the $j$th site split.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch, with $t=\{p,p,p,p,p\}$.
The probabilities of observing ancestral splits for $\siteSplit_1=\emptyset$ are
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_1(\tau, t)=\{\emptyset\}$.
Then, taking $\xi_1\in\ancestralSplitPartition_1(\tau, t)$ we have
$$
\Pr(\ancestralSplitRV=\xi_1 \mid \siteSplitRV=\emptyset, \tau, t) =
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_5=\{1,2\}$ we have
$$
\Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3,
$$
$$
\Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
\Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^3(1-p)^2.
$$
Here, the set of most likely ancestral states is $\ancestralSplitPartition_5(\tau, t)=\{\emptyset,\{1,2\}\}$, and, for $\xi_5\in\ancestralSplitPartition_5(\tau, t)$,
$$
\Pr(\ancestralSplitRV=\xi_5 \mid \siteSplitRV=\{1,2\}, \tau, t) =
p^2(1-p)^3.
$$

@@

To compute the likelihood of observing a set of data, we need $\Pr(\ancestralStateColumn=\ancestralSplit_k\mid \alignmentColumn=\siteSplit_j,\tau,t)$ for each $\ancestralSplit_k$ and $\siteSplit_j$.
We remind the reader that the probability of a character change along a branch with fidelity parameter $\theta$ is $(1-\theta)/2$, while the probability of a character remaining the same is $(1+\theta)/2$.
See Fig.~\ref{fig:example_likelihoods} for the parameters on an example site pattern on the Farris tree.
Likelihood computations for all site patterns and ancestral states are in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods}.
Taking maxima row-wise of each table results in Table~\ref{tab:likelihoods}.
